---
title: "Plotting the distribution of taxa"
author: "Sur Herrera Paredes"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: TRUE
---

```{r}
library(tidyverse)
```
 
# Read data

```{r}
Tab <- read_tsv("data/rhizo/otu_table.tsv")
Tab
```


```{r}
Tab <- read_tsv("data/rhizo/otu_table.tsv",
                col_types = cols(otu_id = col_character(),
                                 .default = col_number()))
Tab
```


# Basic barplot 

We need to think back to the original figure and reformat our data to have one
column for the x-axis and another for the y-axis

```{r}
Tab %>%
  pivot_longer(-otu_id, names_to = "sample_id", values_to = "count")
```

Lets create a smaller subset of the data to make some basic plots

```{r}
dat <- Tab %>%
  pivot_longer(-otu_id, names_to = "sample_id", values_to = "count") %>%
  filter(otu_id %in% c("OTU_14834", "OTU_18567", "OTU_14402", "OTU_14822"))
print(dat)
```

Now we create a basic barplot that shows relative abundances of the 4
selected OTUs
```{r}
dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill")
```

Now we re-make the plot with some beautification

```{r}
dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90))
```

# Adding sample metadata

Read sample metadata

```{r}
Meta <- read_tsv("data/rhizo/sample_metadata.tsv",
                 col_types = cols(.default = col_character()))
Meta
```

We **join** our long format dataset with the sample metadata table

```{r}
print(dat)
dat <- dat %>%
  left_join(Meta, by = "sample_id")
dat
```
We add facts with `facet_grid`

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill") +
  facet_grid(~ fraction) +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(color = "black"))
p1
```
We set `scales="free_x"` to plot a different set of samples on each
facet.

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill") +
  facet_grid(~ fraction, scales = "free_x") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(color = "black"))
p1
```

We repeat the plot with some beautification


```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill", width = 1) +
  facet_grid(~ fraction, scales = "free_x") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(color = "black"),
        strip.text = element_text(face = "bold"),
        strip.background = element_blank())
p1
```
We can add an extra variable to the facet

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill", width = 1) +
  facet_grid(~ fraction + soil, scales = "free_x") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(color = "black"),
        strip.text = element_text(face = "bold"),
        strip.background = element_blank())
p1
```

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = otu_id), stat = "identity", position = "fill", width = 1) +
  facet_grid(~ fraction + soil, scales = "free_x", space = "free_x") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(color = "black"),
        strip.text = element_text(face = "bold"),
        strip.background = element_blank())
p1
```

## Excercise 1

Re-make the plot above but using accession instead of soil type in the facets

## Excercise 2

Re-make the plot above but show all three variables (fraction, accession and
soil) in the facets. What happens if you change the order of the terms
in the facet formulas? is any of the orders better?

# Adding OTU taxonomy

```{r}
Tax <- read_tsv("data/rhizo/otu_taxonomy.tsv",
                col_types = cols(otu_id = col_character(),
                                 Phylum = col_character()))
Tax
```
Let's include all OTUs in our data table now

```{r}
dat <-Tab %>%
  pivot_longer(-otu_id, names_to = "sample_id",
               values_to = "count")
print(dat)
```

As we did before with sample metadata, we use `left_join` to add OTU
taxonomic information to our data table

```{r}
dat <- dat %>%
  left_join(Tax, by = "otu_id")
print(dat)
```

Let's make our basic plot, but using `Phylum` instead of `otu_id`

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill", width = 1) +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(face = "bold")) 
p1
```

Can we find better colors

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill", width = 1) +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(face = "bold")) 
p1
```
I want phyla to be ordered by abundance and "unclassified" to be
at the end

Code to get order not shown but results are in `mean_freqs`

```{r, echo=FALSE}
mean_freqs <- dat %>%
  group_by(sample_id, Phylum) %>%
  summarise(phyl_count = sum(count),
            .groups = 'drop') %>%
  group_by(sample_id) %>%
  summarise(phyl_freq = phyl_count / sum(phyl_count),
            Phylum = Phylum,
            .groups = 'drop') %>%
  group_by(Phylum) %>%
  summarise(mean = mean(phyl_freq),
            .groups = 'drop') %>%
  arrange(desc(mean))
print(mean_freqs)
```
Create a vector with the desired order of phyla and use it to convert
phylum column to a factor.

```{r}
phyla_order <- c("Proteobacteria",                                                                                                                      
                 "Actinobacteria",                                                                                                        
                 "Bacteroidetes",                                                                                        
                 "Acidobacteria",                                                                       
                 "Firmicutes",                                                         
                 "Cyanobacteria",                                     
                 "Verrucomicrobia",                 
                 "Gemmatimonadetes",                                             
                 "Armatimonadetes",
                 "Chloroflexi",
                 "unclassified")

dat <- dat %>%
  mutate(Phylum = factor(Phylum, levels = phyla_order))
dat

```


```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill", width = 1) +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 90),
        axis.text.y = element_text(face = "bold")) 
p1
```
## Excercise

Use `scale_color_manual` to manually select a good set of colors for this plot

# Combining sample metadata and OTU taxonomy

We can use `lef_join` twice in a row to get a combinde data table with all
the information

```{r}
dat <- Tab %>%
  pivot_longer(-otu_id, names_to = "sample_id", values_to = "count") %>%
  left_join(Meta, by = "sample_id") %>%
  left_join(Tax, by = "otu_id")
print(dat)
```

To get the same order as above we convert the Phylum column to a factor

```{r}
dat <- dat %>%
  mutate(Phylum = factor(Phylum, levels = phyla_order))
dat
```


Now we can combine facets and phylum color and ordering in just one plot

```{r}
p1 <- dat %>%
  ggplot(aes(x = sample_id, y = count)) +
  facet_grid(~ fraction + soil, scales = "free_x", space = "free_x") +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill", width = 1) +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  scale_fill_brewer(palette = "Paired") +
  theme(axis.text.x = element_text(angle = 90, size = 6),
        axis.text.y = element_text(color = "black"),
        strip.text = element_text(face = "bold"),
        strip.background = element_blank())
p1
```


We save the plot to a file

```{r}
ggsave("rhizo_phylo_distribution.png", p1, width = 8, height = 4)
```



# Extra excercises

Look at the files at [data/hmp_v13](data/hmp_v13) which contain much bigger
data tables generated from the Human Microbiome Project (HMP).

Can you make similar plots illustrating the bacterial taxonomic distributions
of the Stool and Saliva? Are there any differences by sex?

Can you order the samples by Proteobacteria relative abundance in the
rhizosphere dataset? what about the Bacteroidetes abundance in the
HMP dataset? (HINT: Use the same approach we used to sort taxonomic groups)


# Session info

```{r}
sessionInfo()
```

